---
title: "Functions, loops, conditionals and classes"
author: "Billy Langlet"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
```

## Functions

The simplest R function does not provide default values for arguments, include a return statement or provide argument names in the function call.

```{r simple function anatomy}
my_func <- function(x, y) {
  z <- x * y
}

obj <- my_func(10, 5)
obj
```

There are several additional features for functions.

#### Scope

Usually object created in functions are destroyed when the function ends. However, if you use the "\<\<-" assignment operator the object will be declared "global", i.e., accessible outside of the function.

```{r function scope}
public_func <- function(x, y) {
  z <<- x * y
}

obj <- public_func(3, 7)
```

#### Default arguments

Many functions contain default values for arguments that are often a certain value.

```{r default arguments in functions}
default_func <- function(x = 1, y = 5) {
  z <- x * y
}

obj <- default_func()
obj
obj <- default_func(10, 5)
obj
```

#### Closures

A closure is a function that retains access to variables from the enclosing (outside) function after execution of outside function.

```{r closure functions}
outer_function <- function(x) { 
  inner_function <- function(y) { 
    return(x + y)    
  }
  return(inner_function)
}

closure <- outer_function(10)
closure
result <- closure(5)
result
result2 <- closure(10)
result2
```

#### Anonymous functions

Anonymous functions are usually used when there is no need for reuse but other functions require a function as an argument.

```{r anonymous functions}
function(x) {x * 5}

# Example
sapply(mtcars, function(x) {x * 5})
```

#### Ellipses

Ellipses (...) allow functions to take an unspecified number of arguments

```{r ellipses in functions}
ellipse_func <- function(x, y, ...) {
  z <- x + y

  if (length(list(...)) > 0) {
    z <- z + sum(...)
  }

  return(z)
}

ellipse_func(1, 5)
ellipse_func(1, 2, 3, 4, 5, 6, 7, 8)
ellipse_func(... = c(5, 6, 7), y = 5, x = 2)
```

#### Recursion

A function can call itself.

```{r recursion in functions}
factorial <- function(n){
  if (n == 0 | n == 1) {
    return(1)
  } else {
    return(n * factorial(n - 1))
  }
}

factorial(10)
```

#### Optional documentation

```{r optional documentation in functions}
# This will not create the wanted behaviour, since we have not created a package yet

#' Add together two numbers
#'
#' @param x A number or vector
#' @param y A number or vector
#'
#' @return A number of vector
#' @examples
#' add(1, 1)
add <- function(x, y) {
  x + y
}
```

#### Higher-order functions

Higher order functions either take a function as an argument or returns a function.

```{r higher-order functions}
numbers <- list(1, 4, 9, 16)

square_roots <- lapply(numbers, sqrt)
```

## Loops

#### For loop

The for loop is a control variable that tracks the current position in a sequence.

```{r for loop anatomy}
# Alternative 1
for (i in 1:10) {
  print(i)
}

my_vector <- c(1, 5, 2, 12)

# Alternative 2
for (i in my_vector) {
  print(i)
}
```

#### While loop

The while loop is a control variable that evaluates to a logical value from the comparison between two values.

```{r while loop anatomy}
i = 1
while(i < 10) {
  print(i)
  i <- i + 1
}
```

#### Break

Terminates a loop prematurely

```{r break}
df <- list(1, 2, 3)
df2 <- list(mtcars)

break_func <- function(data) {
  for (i in data) {
    if (is.null(dim(i))) {
      print("Wrong number of dimensions")
      break
    }
    print(data[[1]][1, 1])
  }
}

break_func(df)
break_func(df2)
```

#### Return

You can use return to exit a function, and as an extension, any loop within the function.

```{r return}
return_func <- function(x, y) {
  for (i in x) {
    print(i)
    if (i == y) {
      return(paste0("The letter ", y, " is present in the vector"))
    }
  }
  return(-1)
}

return_func(letters, "g")
return_func(letters, "10")
```

#### Next

Skips the current itteration.

```{r next}
skip_func <- function(x, y) {
  tmp_data <- c()
  for (i in x) {
    if (class(i)  == "character") {
      next
    }
    tmp_data <- c(tmp_data, i * y)
  }
  return(tmp_data)
}

skip_func(list(1, "a", 5, 10, "b"), 5)
```

## Conditionals

```{r conditional anatomy}
cond_func <- function(x){
  if (x < 5) {
    y <- "less than"
  } else if (x > 5) {
    y <- "more than"
  } else {
    y <- "equal"
  }
  return(y)
}

cond_func(1)
cond_func(10)
```

#### Multiple conditions

```{r multiple conditions}
multi_func <- function(BMI) {
  if (BMI < 30 && BMI > 18.5) {
    x <- "no health check"
  } else {
    x <- "health check"
  }
  return(x)
}

multi_func(20)
```

#### Guard clause

To prevent a function from running if something is "wrong".

```{r guard clause}

guard_func <- function (x) {
  if (class(x) != "list") {
    return("argument X is not a list")
  }
  # Some operation
}

guard_func(list(1, 2, 3))
guard_func(mtcars)
```

#### Default values

```{r default values}

default_func <- function(x) {
  if (class(x) == "list") {
    y <- x[[1]]
  } else {
    y <- -1  
  }
  return(y)
}

default_func(list(1, 2, 3))
default_func(mtcars)
```

#### Try catch

There is also a function to create "Try catch" expressions.

```{r try catch}
evaluate_div <- function(x) {
  tryCatch(
    expr = {
      message("Successfully executed the log(x) call.")
      print(log(x))
    },
    error = function(e){
      message('Caught an error!')
      print(e)
    },
    warning = function(w){
      message('Caught an warning!')
      print(w)
    },
    finally = {
      message('All done, quitting.')
    })
}

evaluate_div(5)
evaluate_div(-10)
evaluate_div("Hello")
```

## Classes

Basics of classes in R

```{r classes}
# S3 class (not declared value type of arguments)
obj <- c(x = 3, y = 4)
class(obj) <- "Point2D"

# S4 class (declared value type of arguments)
setClass("Point2D", slots = c(x = "numeric", y = "numeric"))

point1 <- new("Point2D", x = 3, y = 4)
point2 <- new("Point2D", x = 0, y = 10)

# Other declaration of S4 class
myClass <- setClass("Point3D", slots = c(x = "numeric", y = "numeric", z = "numeric"))

point3 <- myClass(x = 1, y = 2, z = 3)
point4 <- myClass(x = 1, y = 2, z = 5)
```

#### Extending class

Adding a method (function) to a class can be done in R.

```{r extending class}
distance.Point2D <- function(p1, p2) {
  sqrt((p1@x - p2@x)^2 + (p1@y - p2@y)^2)
}

print.Point2D <- function(p) {
  cat("Point(", p@x, ",", p@y, ")\n")
}

distance.Point2D(point1, point2)
print.Point2D(point1)

# There are additional ways to handle more complex classes and methods using setMethod() etc
```
