---
title: "Algorithms"
author: "Billy"
date: "2023-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Algorithms

## Big O notation

Conducting a Big O notation analysis involves evaluating the time or space complexity of an algorithm to understand how it scales with input size.

#### Example O(1)

```{r O(1)}
O1_func <- function(arr) {
  return(arr[1])
}

O1_func(1:10000)
```

#### Example O(log n) - Binary search

```{r O(log n)}
Ologn_func <- function(arr, target) {
  left <- 1
  right <- length(arr)

  while (left <= right) {
    mid <- left + floor((right - left) / 2)  # Calculate the middle index

    if (arr[mid] == target) {
      return(mid)  # Target found
    } else if (arr[mid] < target) {
      left <- mid + 1  # Search the right half
    } else {
      right <- mid - 1  # Search the left half
    }
  }

  return(-1)  # Target not found
}

Ologn_func(1:10000, 300)
```

#### Example O(n) - Linear search

```{r O(n)}
On_func <- function(arr, target) {
  for (i in arr) {
    if (arr[i] == target) {
      return(target)
    }
  }
  
  return(-1)
}

On_func(1:10000, 10000)
```

#### Example O(n\^2) - Bubble sort

```{r O(n^2)}
On2_func <- function(arr) {
  n <- length(arr)
  for (i in 1:(n - 1)) {
    for (j in 1:(n - i)) {
      if (arr[j] > arr[j + 1]) {
        # Swap elements arr[j] and arr[j + 1]
        temp <- arr[j]
        arr[j] <- arr[j + 1]
        arr[j + 1] <- temp
      }
    }
  }
  return(arr)
}

sorted_arr <- On2_func(sample(1:10000))
```

## Basic algorithms

Some of the more basic algorithms are **searching** and **sorting** algorithms.

#### Example bubble sort

```{r bubble sort}
bubble_sort <- function(arr) {
  if (!is.null(dim(arr))) {
    stop('Not a vector')
  }
  
  n <- length(arr)
  for (i in 1:(n - 1)) {
    for (j in 1:(n - i)) {
      if (arr[j] > arr[j + 1]) {
        # Swap elements arr[j] and arr[j + 1]
        temp <- arr[j]
        arr[j] <- arr[j + 1]
        arr[j + 1] <- temp
      }
    }
  }
  return(arr)
}

sorted_arr <- bubble_sort(sample(1:10000))
```

#### Example insertion sort

```{r insertion sort}
insertion_sort <- function(arr) {
  if (!is.null(dim(arr))) {
    stop('Not a vector')
  }

  n <- length(arr)
  
  for (i in 2:n) {
    key <- arr[i]
    j <- i - 1
    
    # Move elements of arr[1..i-1] that are greater than key
    # to one position ahead of their current position
    while (j >= 1 && arr[j] > key) {
      arr[j + 1] <- arr[j]
      j <- j - 1
    }
    arr[j + 1] <- key
  }
  
  return(arr)
}

insertion_sort(sample(1:10))
insertion_sort(sample(1:10000))
```

#### Example linear search

```{r linear search}
linear_search <- function(arr, target) {
  if (!is.null(dim(arr))) { # we can use the stop statement to exit the function
    stop('Not a vector')
  }
  
  for (i in arr) {
    if (arr[i] == target) {
      return(T)
    }
  }
  return(F)
}

linear_search(1:10000, 7000)
```

### Example binary search

```{r binary search}
binary_search <- function(arr, target) {
  if (!is.null(dim(arr))) { # we can use the stop statement to exit the function
    stop('Not a vector')
  }
  
  left <- 1
  right <- length(arr)

  while (left <= right) {
    mid <- left + floor((right - left) / 2)  # Calculate the middle index

    if (arr[mid] == target) {
      return(mid)  # Target found
    } else if (arr[mid] < target) {
      left <- mid + 1  # Search the right half
    } else {
      right <- mid - 1  # Search the left half
    }
  }
  return(-1)  # Target not found
}

Ologn_func(1:10000, 300)
```

## More complex algorithms

Divide and conquer...

#### Example quick sort

```{r quick sort}
quick_sort <- function(arr) {
  if (length(arr) <= 1) {
    return(arr)
  }
  
  pivot <- arr[length(arr) %/% 2]  # Choose a pivot element (middle of the array)
  left <- arr[arr < pivot]         # Elements smaller than the pivot
  middle <- arr[arr == pivot]      # Elements equal to the pivot
  right <- arr[arr > pivot]        # Elements greater than the pivot
  
  return(c(quick_sort(left), middle, quick_sort(right)))
}

quick_sort(sample(1:10000))
```

#### Example Karatsuba algoritm

```{r Karatsuba algorithm}
karatsuba <- function(x, y) {
  if (x < 10 || y < 10) {
    return(x * y)
  }
  
  # Calculate the number of digits in both numbers
  n <- max(nchar(x), nchar(y))
  n2 <- n %/% 2

  # Split the numbers into two halves
  a <- x %/% 10^n2
  b <- x %% 10^n2
  c <- y %/% 10^n2
  d <- y %% 10^n2

  # Recursively calculate the three products required
  ac <- karatsuba(a, c)
  bd <- karatsuba(b, d)
  ad_bc <- karatsuba((a + b), (c + d)) - ac - bd

  # Combine the products to get the final result
  result <- ac * 10^(2 * n2) + ad_bc * 10^n2 + bd

  return(result)
}

karatsuba(124578, 326598)
```

## Advanced algorithms

Dynamic programming...

#### Example Fibonacci sequence without dynamic programming

```{r Fibonacci sequence recursive calculation}
fibonacci_recursive <- function(n) {
  if (n <= 0) {
    return(NULL)
  } else if (n == 1) {
    return(0)
  } else if (n == 2) {
    return(c(0, 1))
  } else {
    fib_sequence <- fibonacci_recursive(n - 1)
    next_fib <- tail(fib_sequence, 1) + tail(head(fib_sequence, -1), 1)
    return(c(fib_sequence, next_fib))
  }
}

# Solve the fibonacci sequence
fibonacci_recursive(10)
```

#### Example Fibonacci sequence with dynamic programming

```{r Fibonacci sequence with dynamic programming}
fibonacci_dynamic <- function(n) {
  if (n <= 0) {
    return(NULL)
  } else if (n == 1) {
    return(0)
  } else if (n == 2) {
    return(c(0, 1))
  } else {
    fib_sequence <- numeric(n)
    fib_sequence[1] <- 0
    fib_sequence[2] <- 1
    for (i in 3:n) {
      fib_sequence[i] <- fib_sequence[i - 1] + fib_sequence[i - 2]
    }
    return(fib_sequence)
  }
}

# Solve the fibonacci sequence
fibonacci_dynamic(10)
```

#### Example 0/1 Knapsack problem

0/1 means you can either take or leave the item. The other version is the fractional knapsack problem, where you can take fractions of items.

```{r Knapsack problem}
knapsack_01 <- function(values, weights, capacity) {
  n <- length(values)
  dp <- matrix(0, nrow = n + 1, ncol = capacity + 1)

  for (i in 1:n) {
    for (w in 1:capacity) {
      if (weights[i] <= w) {
        dp[i + 1, w] <- max(dp[i, w], dp[i, w - weights[i]] + values[i])
      } else {
        dp[i + 1, w] <- dp[i, w]
      }
    }
  }

  selected_items <- integer(0)
  i <- n
  w <- capacity

  while (i > 0 && w > 0) {
    if (dp[i + 1, w] != dp[i, w]) {
      selected_items <- c(selected_items, i)
      w <- w - weights[i]
    }
    i <- i - 1
  }

  return(list(max_value = dp[n + 1, capacity], selected_items = selected_items))
}

# Solve the 0/1 Knapsack Problem
knapsack_01(c(40, 100, 120), c(10, 20, 30), 50)
```

Greedy algorithms...

#### Example Fractional knapsack problem

```{r Fractional knapsack problem}
fractional_knapsack <- function(values, weights, capacity) {
  # Calculate the value-to-weight ratio for each item
  value_per_weight <- values / weights
  
  # Create an index vector to keep track of items
  n <- length(weights)
  item_indices <- 1:n
  
  # Sort items by the value-to-weight ratio in descending order
  sorted_items <- item_indices[order(-value_per_weight)]
  
  # Initialize variables
  total_value <- 0
  knapsack <- rep(0, n)
  current_weight <- 0
  
  # Fill the knapsack greedily
  for (i in sorted_items) {
    if (current_weight + weights[i] <= capacity) {
      # Add the whole item to the knapsack
      knapsack[i] <- 1
      total_value <- total_value + values[i]
      current_weight <- current_weight + weights[i]
    } else {
      # Add a fraction of the item to fill the knapsack to its capacity
      fraction <- (capacity - current_weight) / weights[i]
      knapsack[i] <- fraction
      total_value <- total_value + fraction * values[i]
      break  # Knapsack is full
    }
  }
  
  # Return the selected items and total value
  result <- list(
    selected_items = knapsack,
    total_value = total_value
  )
  return(result)
}

# Solve the Fractional Knapsack Problem
fractional_knapsack(c(40, 100, 120), c(10, 20, 30), 25)
```

#### Example Dijkstra's shortest path

```{r Dijikstras shortest path}
# Function to find the vertex with the smallest tentative distance.
find_min_distance_vertex <- function(distances, visited) {
  min_distance <- Inf
  min_vertex <- NULL
  
  for (vertex in names(distances)) {
    if (!visited[vertex] && distances[vertex] < min_distance) {
      min_distance <- distances[vertex]
      min_vertex <- vertex
    }
  }
  return(min_vertex)
}

# Dijkstra's algorithm
dijkstra <- function(graph, start) {
  # Initialize distances to all nodes as infinity and the start node as 0.
  distances <- rep(Inf, length(graph))
  names(distances) <- names(graph)
  distances[start] <- 0
  
  # Create a boolean array to keep track of visited nodes.
  visited <- rep(FALSE, length(graph))
  names(visited) <- names(graph)
  
  for (i in 1:length(graph)) {
    # Find the node with the smallest tentative distance.
    current_vertex <- find_min_distance_vertex(distances, visited)
    
    # Mark the current node as visited.
    visited[current_vertex] <- TRUE
    
    # Update distances to neighboring nodes.
    for (neighbor in names(graph[[current_vertex]])) {
      if (!visited[neighbor]) {
        new_distance <- distances[current_vertex] +
          graph[[current_vertex]][[neighbor]]
        if (new_distance < distances[neighbor]) {
          distances[neighbor] <- new_distance
        }
      }
    }
  }
  return(distances)
}

# Example graph represented as an adjacency list.
graph <- list(
  'A' = list('B' = 2, 'C' = 6),
  'B' = list('A' = 2, 'D' = 5),
  'C' = list('A' = 6, 'D' = 8),
  'D' = list('B' = 5, 'C' = 8, 'E' = 10, 'F' = 15),
  'E' = list('D' = 10, 'F' = 6, 'G' = 2),
  'F' = list('D' = 15, 'E' = 6, 'G' = 6),
  'G' = list('E' = 2, 'F' = 6)
)

# Solve the shortest distance
shortest_distances <- dijkstra(graph, 'A')
```
